Estimation of Manning's Friction Coefficient using DART Buoys Data during \tohoku Tsunami

### 2013-12-24 paper/tsunami.tex
more editing to the intro and other edits

Manning's $n$ law provides a relationship between the friction coefficient appearing in the shallow
water equations governing tsunamis and the roughness parameter $n$. 
This Manning's $n$ coefficient is variable that is highly spatial 
and dependent on the surface characteristics of the sea-floor. 
The direct measurement of $n$ is therefore impossible, instead, values
are often assigned from tables that are empirically estimated
for common types of land that are determined from satellite images.
As a result, the Manning's $n$ coefficients used in tsunami models
often contain large amounts of uncertainty.

The uncertainty in the Manning's $n$ coefficient leads to uncertainties
in the predicted quantities of interest such as the water elevations
and becomes a health risk during tsunami events. In addition, 
the prediction of these quantities of interest are often highly sensitive to 
the changes in these uncertain parameters. 
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the roughness parameter $n$. To this end,
we characterize the roughness $n$ by three different parameters
that are varied in three different regions but are assumed
constant in each region. The three regions are as follows: 
on-shore whose coefficient is denoted by $n_1$, near-shore 
denoted by $n_2$, and deep-water denoted by $n_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $n$ coefficients a uniform 
distribution was assumed as the prior where $n$ was sampled in the interval $n \in [0.005,0.2]$.

\section{The \tohoku Tsunami} 
\label{ssub:tohoku}

@@

\begin{equation}
    C_f = \frac{g n^2}{h^{5/3}}
\end{equation}
where $n$ is an empirically determined parameter.

@@

\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
at the gauges ready.... Not sure if this explanation is worth including here ///}

@@

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right).

>>>>

Manning's $n$ law provides a relationship between the friction coefficient 
(Equation \ref{eq:coef}) appearing in the shallow
water equations governing tsunamis (see Equation \ref{eq:swe})
and the roughness parameter $n$. 
This Manning's $n$ coefficient is variable that is highly spatial 
and dependent on the surface characteristics of the sea-floor. 
The direct measurement of $n$ is therefore impossible, instead, values
are often assigned from tables that are empirically estimated
for common types of land that are determined from satellite images.
As a result, the Manning's $n$ coefficients used in tsunami models
often contain large amounts of uncertainty.

The uncertainty in the Manning's $n$ coefficient leads to uncertainties
in the predicted quantities of interest such as the water elevations
and becomes a health risk during tsunami events. In addition, 
the prediction of these quantities of interest are often highly sensitive to 
the changes in these uncertain parameters. 
In this work, we seek to quantify parametric uncertainty namely due 
to the uncertainty in the friction parameter $n$. To this end,
we characterize the Manning's $n$ coefficient by three different parameters
that are varied in three different regions but are assumed
constant in each region. The three regions are as follows: 
on-shore whose coefficient is denoted by $n_1$, near-shore 
denoted by $n_2$, and deep-water denoted by $n_3$.
The three regions are shown in Figure~\ref{fig:ceofs}.
To carry out the uncertainty quantification we have employed polynomial chaos expansions 
as used in \cite{sraj:2013a,sraj:2013b}. The method and its application
to our problem is described in Section~\ref{sec:uqpce}.  
To represent the uncertainty in the Manning's $n$ coefficients a uniform 
distribution was assumed as the prior where $n$ was sampled in the interval $n \in [0.005,0.2]$.

\section{The \tohoku Tsunami} 
\label{sec:tohoku}

@@

\begin{equation}
    C_f = \frac{g n^2}{h^{5/3}}
\label{eq:coef}
\end{equation}
where $n$ is an empirically determined parameter.

@@

%\alert{Need to describe how we handled the AMR in the UQ calculation: Ihab: Nothing special, we had only four time
%steps so I made sure that we have the same AMR grid for the 125 ensemble of realizations. It turned out
%one time step results in different AMR grid for some realizations so I ignored it and didnt include in the discussion.
%Regarding the inverse problem and estimating N, there was no need to worry since I had the local data
%at the gauges ready.... Not sure if this explanation is worth including here ///}

@@

The observations used in this study are from the Deep-ocean Assessment and Reporting of Tsunamis (DART) buoy system developed and maintained by the National Oceanic and Atmospheric Administration (NOAA).  The purpose of the network is to provide early-warning detection and forecasting of tsunami propagation in the Pacific Ocean \alert{need citation for this?}.  The DART buoys closest to the earthquake source of the \tohoku tsunami were buoys 21401, 21413, 21418, and 21419 whose locations are shown in Figure~\ref{fig:setup}(left) and whose de-tided data for the event are shown in Figure~\ref{fig:setup}(right). \comment{maybe more description of the data in the figure}.
